---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# disprofas

<!-- badges: start -->
<!-- badges: end -->

Functions for the determination of similarity of highly variable dissolution profiles of two drug formulations are provided. Estimation of *f*~1~, *f*~2~ and bootstrap *f*~2~ is implemented. In addition, functions following the model-independent MCR (multivariate confidence region) procedure and the *T*^2^-test for equivalence procedure are available.

## Installation

**So far, the package has not yet been submitted to CRAN.** The development version is available from [GitHub](https://github.com/piusdahinden/disprofas). Installation from this source is easily done by using `install_github()` function from the `devtools` package:

``` r
# install.packages("devtools")
devtools::install_github("piusdahinden/disprofas")
```

## Examples

### Example 1
*Example 1* illustrates how to solve a common problem by aid of the bootstrap *f*~2~ procedure proposed by [Shah et al. (1998)](https://doi.org/10.1023/A:1011976615750) using a data set containing the dissolution data of one reference batch and one test batch of *n* = 12 tablets each, i.e. the dissolution profiles of the % drug release observed at 0, 30, 60, 90 and 180 minutes (See Shah et al. (1998), Table 4).

``` {r example_1, message = FALSE}
library(disprofas)

# Data frame
str(dip2)

# Perform estimation and print a summary
res1 <- bootstrap_f2(data = dip2[dip2$batch %in% c("b0", "b4"), ],
                         tcol = 5:8, grouping = "batch",
                         R = 200, new_seed = 421, useEMA = "no")

class(res1)
summary(res1)

# Prepare graphical representation
plot(res1)
```


### Example 2
*Example 2* illustrates how to solve a common problem by aid of the model-independent non-parametric multivariate confidence region (MCR) procedure proposed by [Tsong et al. (1996)](https://doi.org/10.1177/009286159603000427) using a data set containing the dissolution data of one reference batch and one test batch of *n* = 6 tablets each, i.e. the dissolution profiles of the % drug release observed at 5, 10, 15, 20, 30, 60, 90 and 120 minutes (see Tsong et al. (1996), Table 1).

``` {r example_2, message = FALSE}
library(disprofas)

# Data frame
str(dip3)

# Perform estimation and print a summary
res2 <- mimcr(data = dip3, tcol = 4:6, grouping = "batch")

class(res2)
summary(res2)
```


### Example 3
*Example 3* illustrates how to solve a common problem by aid of the *T*^2^-test for equivalence procedure proposed by [Hoffelder (2016)](https://www.ecv.de/suse_item.php?suseId=Z|pi|8430) using a data set containing the dissolution data of one reference batch and one test batch of *n* = 12 capsules each, i.e. the dissolution profiles of the % drug release observed at 15, 20 and 25 minutes (see Hoffelder (2016), Figure 1 (data not shown in publication, but the data set is available on CRAN, package [T2EQ](https://CRAN.R-project.org/package=T2EQ), data set `ex_data_pharmind`)).

``` {r example_3, message = FALSE}
library(disprofas)

# Data frame
str(dip4)

# Perform estimation and print a summary
res3 <- mimcr(data = dip4, tcol = 2:4, grouping = "type")
summary(res3)
```


### Example 4
*Example 4* illustrates tolerance interval approach proposed by [Martinez & Zhao (2018)](https://doi.org/10.1208/s12248-018-0238-1) using the data set that was used in *Example 1*.

``` {r example_4, message = FALSE}
library(disprofas)

# Data frame
str(dip1)

# Perform estimation and print a summary
res4 <- mztia(data = dip1, shape = "wide", tcol = 3:10, grouping = "type",
                reference = "R")

class(res4)
summary(res4)

# Prepare graphical representation
ggres4 <- plot_mztia(res4)
class(ggres4)
plot(ggres4)
```


## Contact
[Pius Dahinden](https://github.com/piusdahinden), Tillotts Pharma AG


---
